---
title: "Comparaison packages boral et jSDM"
author: "Jeanne Clément"
date: "23/08/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 7, fig.height = 7,
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  highlight = TRUE
)
```

# Comparison between packages boral and jSDM 

## Simulated dataset :300 sites and 100 species

### Fit with boral

```{r simulation-dataset, echo=F, include=F}

# Toy example : 
nsp <- 100
nsite <- 300
nl <- 2
seed <- 1234
set.seed(seed)

# Ecological process (suitability)
x1 <- rnorm(nsite,0,1)
x2 <- rnorm(nsite,0,1)
X <- cbind(rep(1,nsite),x1,x2)
np <- ncol(X)

# Simulation of parameters
W <- cbind(rnorm(nsite,0,1),rnorm(nsite,0,1))
data=cbind (X,W)
beta.target <- t(matrix(runif(nsp*np,-2,2), byrow=TRUE, nrow=nsp))
l.zero <- 0
l.diag <- runif(2,0,2)
l.other <- runif(nsp*2-3,-2,2)
lambda.target <- t(matrix(c(l.diag[1],l.zero,l.other[1],l.diag[2],l.other[-1]), byrow=T, nrow=nsp))
param.target <- rbind(beta.target,lambda.target)
param.target=rbind(beta.target,lambda.target)
Valpha.target=0.5
V=1
alpha.target <- rnorm(nsite,0,sqrt(Valpha.target))

# Simulation of response data with probit link
probit_theta=X %*% beta.target + W %*% lambda.target + alpha.target
e=matrix(rnorm(nsp*nsite,0,sqrt(V)),nsite,nsp)
Z_true <- probit_theta + e
Y=matrix (NA, nsite,nsp)
for (i in 1:nsite){
  for (j in 1:nsp){
    if ( Z_true[i,j] > 0) {Y[i,j] <- 1}
    else {Y[i,j] <- 0}
  }
}
```

```{r boral-simulation, cache=TRUE, echo=F, include=F}
library(boral)
T1<-Sys.time() 
mod_boral_sim <- boral(y=Y, X=X[,-1], lv.control=list(num.lv=nl, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nsite),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=30000, n.iteration=35000,n.thin=5,seed=123))
T2<-Sys.time() 
T_boral_sim=difftime(T2, T1,units="mins")

# Predicted probit theta 
probit_theta_pred_sim=mod_boral_sim$row.coefs[[1]]$mean + X[,-1] %*% t(mod_boral_sim$X.coefs.mean)  + mod_boral_sim$lv.coefs.mean[,"beta0"] + mod_boral_sim$lv.mean%*%t(mod_boral_sim$lv.coefs.mean[,-1])
# RMSE
SE=(probit_theta-probit_theta_pred_sim)^2
RMSE_boral_sim=sqrt(sum(SE/(nsite*nsp)))
# Deviance 
logL=0
for (i in 1:nsite){
  for (j in 1:nsp){
    theta <- pnorm(probit_theta_pred_sim[i,j])
    logL=logL + dbinom(Y[i,j],1,theta,1)  
  }
}
Deviance_boral_sim <- -2*logL
```

### Fit with jSDM 
```{r jSDM-simulation, cache=TRUE, echo=F, include=F}
library(jSDM)

# Fit the model
T1<-Sys.time() 
mod_jSDM_block_sim <-  jSDM::jSDM_probit_block (
  presence_site_sp=Y, 
  site_suitability=~.,   
  site_data=as.data.frame(X[,-1]), n_latent=2,
  burnin=30000, mcmc=5000, thin=5,
  alpha_start=0, beta_start=0, lambda_start=0, W_start=0,
  V_alpha_start=1, shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  seed=123, verbose=1)
T2<-Sys.time() 
T_jSDM_block_sim=difftime(T2, T1,units="mins")

# RMSE
SE=(probit_theta-mod_jSDM_block_sim$probit_theta_pred)^2
RMSE_jSDM_block_sim=sqrt(sum(SE/(nsite*nsp)))
```

### Comparison of estimated parameters 

```{r jSDM-boral-simulation, echo=F}
# Alpha et V_alpha
par(mfrow=c(1,1))
plot(mod_boral_sim$row.coefs[[1]]$mean,summary(mod_jSDM_block_sim$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
points(mod_boral_sim$row.sigma[[1]]$mean,summary(mod_jSDM_block_sim$mcmc.Valpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)
# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  for (p in 1:np){
    jSDM_betas[j,p] <- mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,p])
  }
}
boral_betas <- cbind(mod_boral_sim$lv.coefs.mean[,"beta0"],mod_boral_sim$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by boral", ylab="betas estimated by jSDM", main="Species fixed effect beta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  for (l in 1:nl){
    jSDM_lambdas[j,l] <- mean(mod_jSDM_block_sim$mcmc.sp[[paste0("sp_",j)]][,np+l])
  }
}
boral_lambdas <- mod_boral_sim$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda on latent variables", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- colMeans(mod_jSDM_block_sim$mcmc.latent[[paste0("lv_",l)]])
}
plot(mod_boral_sim$lv.mean, jSDM_lvs, xlab="latent variables estimated by boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_sim, mod_jSDM_block_sim$probit_theta_pred, xlab="probit theta estimated by Boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

par(mfrow=c(1,1))
plot(W[,2],jSDM_lvs[,2], main="Latent variable W_1",xlab="observed", ylab="fitted",cex.lab=1.5,cex.main=1.5,font.lab=2)
points(W[,2],mod_boral_sim$lv.mean[,2], col ='blue')
abline(a=0,b=1,col="red")
legend("bottomright", legend=c("jSDM","boral"), pch = 1, col =c('black','blue'),pt.cex =2, cex=1.2)

```

## Mosquito dataset : 167 sites and 16 species 

### Fit with boral
```{r boral-mosquito, cache=TRUE, echo=F, include=F}
# Import center and reduce Mosquito dataset
Env_mosquito <- read.csv("~/Code/data/Mosquito_Covar.csv", sep=",", header=T)
Env_mosquito <- cbind(scale(Env_mosquito[,1:4]), Env_mosquito[,5:13])
PA_mosquito <- read.csv("~/Code/data/Mosquito_PA.csv", sep=",", header=T)

# Fit the model 
T1 <- Sys.time()
mod_boral_Mosquito <- boral(y=PA_mosquito, X=Env_mosquito, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_mosquito)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=30000, n.iteration=35000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Mosquito <- difftime(T2,T1,units="mins")

# Predicted probit theta 
probit_theta_pred_Mosquito <- mod_boral_Mosquito$row.coefs[[1]]$mean + as.matrix(Env_mosquito) %*% t(mod_boral_Mosquito$X.coefs.mean)  + mod_boral_Mosquito$lv.coefs.mean[,"beta0"] + mod_boral_Mosquito$lv.mean%*%t(mod_boral_Mosquito$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_mosquito)){
  for (j in 1:ncol(PA_mosquito)){
    theta <- pnorm(probit_theta_pred_Mosquito[i,j])
    logL=logL + dbinom(PA_mosquito[i,j],1,theta,1)  
  }
}
Deviance_boral_Mosquito <- -2*logL
```

### Fit with jSDM

```{r jSDM-mosquito, cache=TRUE, echo=F, include=F}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Mosquito <- jSDM_probit_block (
  presence_site_sp=as.matrix(PA_mosquito), 
  site_suitability=~.,   
  site_data=Env_mosquito, n_latent=2,
  burnin=30000, mcmc=5000, thin=5,
  alpha_start=0, beta_start=0, lambda_start=0, W_start=0,
  V_alpha_start=1, shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  seed=123, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Mosquito <- difftime(T2,T1,units="mins")
```

### Comparison of estimated parameters 
```{r jSDM-boral-mosquito, echo=F}
print(paste(nrow(PA_mosquito),"sites and ",ncol(PA_mosquito)," species"), quote=F)
nsp <- ncol(mod_jSDM_block_Mosquito$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Mosquito$model_spec$presences)
nl <- mod_jSDM_block_Mosquito$model_spec$n_latent
np <- nrow(mod_jSDM_block_Mosquito$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,1))
plot(mod_boral_Mosquito$row.coefs[[1]]$mean,summary(mod_jSDM_block_Mosquito$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by boral", ylab="alphas estimated by jSDM", main="Site random effect alpha", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
points(mod_boral_Mosquito$row.sigma[[1]]$mean,summary(mod_jSDM_block_Mosquito$mcmc.Valpha)[[1]]["Mean"], pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  jSDM_betas[j,] <- summary(mod_jSDM_block_Mosquito$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Mosquito$lv.coefs.mean[,"beta0"],mod_boral_Mosquito$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by boral", ylab="betas estimated by jSDM", main="Species fixed effect beta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  jSDM_lambdas[j,] <- summary(mod_jSDM_block_Mosquito$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Mosquito$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda on latent variables", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Mosquito$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Mosquito$lv.mean, jSDM_lvs, xlab="latent variables estimated by boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
# Predictions 
plot(probit_theta_pred_Mosquito, mod_jSDM_block_Mosquito$probit_theta_pred, xlab="probit theta estimated by boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
```

## Eucalypts dataset : 458 sites and 12 species

### Fit with boral 
```{r boral-eucalypts, cache=TRUE, echo=F, include=F}
# Import center and reduce Eucalypts dataset
Env_Eucalypts <- read.csv("~/Code/data/Eucalypts_Covar.csv", sep=",", header=T)
Env_Eucalypts <- cbind(scale(Env_Eucalypts[,c("Rockiness","VallyBotFlat","PPTann", "cvTemp","T0")]),Env_Eucalypts[,c("Sandiness","Loaminess")])
colnames(Env_Eucalypts) <-c("VallyBotFlat","PPTann", "cvTemp","T0","Rockiness","Sandiness","Loaminess")
PA_Eucalypts<- read.csv("~/Code/data/Eucalypts_PA.csv", sep=",", header=T)

# Fit the model 
T1 <- Sys.time()
mod_boral_Eucalypts <- boral(y=PA_Eucalypts, X=Env_Eucalypts, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Eucalypts)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=30000, n.iteration=35000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Eucalypts <- difftime(T2,T1,units="mins")

# Predicted probit theta 
probit_theta_pred_Eucalypts=mod_boral_Eucalypts$row.coefs[[1]]$mean + as.matrix(Env_Eucalypts) %*% t(mod_boral_Eucalypts$X.coefs.mean)  + mod_boral_Eucalypts$lv.coefs.mean[,"beta0"] + mod_boral_Eucalypts$lv.mean%*%t(mod_boral_Eucalypts$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Eucalypts)){
  for (j in 1:ncol(PA_Eucalypts)){
    theta <- pnorm(probit_theta_pred_Eucalypts[i,j])
    logL=logL + dbinom(PA_Eucalypts[i,j],1,theta,1)  
  }
}
Deviance_boral_Eucalypts <- -2*logL
```

### Fit with jSDM 

```{r jSDM-eucalypts, cache=TRUE,echo=F, include=F}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Eucalypts <- jSDM_probit_block (
  presence_site_sp=as.matrix(PA_Eucalypts), 
  site_suitability=~.,   
  site_data=Env_Eucalypts, n_latent=2,
  burnin=30000, mcmc=5000, thin=5,
  alpha_start=0, beta_start=0, lambda_start=0, W_start=0,
  V_alpha_start=1, shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  seed=123, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Eucalypts <- difftime(T2,T1,units="mins")
```

### Comparison of estimated parameters 
```{r jSDM-boral-Eucalypts, echo=F}
nsp <- ncol(mod_jSDM_block_Eucalypts$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Eucalypts$model_spec$presences)
nl <- mod_jSDM_block_Eucalypts$model_spec$n_latent
np <- nrow(mod_jSDM_block_Eucalypts$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=(c(1,1)))
plot(mod_boral_Eucalypts$row.coefs[[1]]$mean,summary(mod_jSDM_block_Eucalypts$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by Boral", ylab="alphas estimated by jSDM", main="Site random effect alpha",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
points(mod_boral_Eucalypts$row.sigma[[1]]$mean,summary(mod_jSDM_block_Eucalypts$mcmc.Valpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  jSDM_betas[j,] <- summary(mod_jSDM_block_Eucalypts$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Eucalypts$lv.coefs.mean[,"beta0"],mod_boral_Eucalypts$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by Boral", ylab="betas estimated by jSDM", main="Species fixed effect beta", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  jSDM_lambdas[j,] <- summary(mod_jSDM_block_Eucalypts$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Eucalypts$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda on latent variables", cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Eucalypts$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Eucalypts$lv.mean, jSDM_lvs, xlab="latent variables estimated by boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_Eucalypts, mod_jSDM_block_Eucalypts$probit_theta_pred, xlab="probit theta estimated by boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
```

## Frogs dataset : 104 sites and 9 species

### Fit with boral 

```{r boral-frogs, cache=TRUE,echo=F, include=F}
# Import center and reduce Frogs dataset
Frogs <- read.csv("~/Code/data/frogs.csv", sep=",", header=T)
Env_Frogs <- cbind(scale(Frogs[,1]),Frogs[,2],scale(Frogs[,3]))
colnames(Env_Frogs) <- colnames (Frogs[,1:3])
PA_Frogs <- Frogs[,4:12]

# Fit the model
T1 <- Sys.time()
mod_boral_Frogs <- boral(y=PA_Frogs, X=Env_Frogs, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Frogs)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=30000, n.iteration=35000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Frogs <- difftime(T2,T1,units="mins")

# Predicted probit theta 
probit_theta_pred_Frogs=mod_boral_Frogs$row.coefs[[1]]$mean + as.matrix(Env_Frogs) %*% t(mod_boral_Frogs$X.coefs.mean)  + mod_boral_Frogs$lv.coefs.mean[,"beta0"] + mod_boral_Frogs$lv.mean%*%t(mod_boral_Frogs$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Frogs)){
  for (j in 1:ncol(PA_Frogs)){
    theta <- pnorm(probit_theta_pred_Frogs[i,j])
    logL=logL + dbinom(PA_Frogs[i,j],1,theta,1)  
  }
}
Deviance_boral_Frogs <- -2*logL
```

### Fit with jSDM 

```{r jSDM-frogs, cache=TRUE,echo=F, include=F}
# Fit the model 
T1 <- Sys.time()
mod_jSDM_block_Frogs <- jSDM_probit_block (
  presence_site_sp=as.matrix(PA_Frogs), 
  site_suitability=~.,   
  site_data=as.data.frame(Env_Frogs), n_latent=2,
  burnin=30000, mcmc=5000, thin=5,
  alpha_start=0, beta_start=0, lambda_start=0, W_start=0,
  V_alpha_start=1, shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  seed=123, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Frogs <- difftime(T2,T1,units="mins")
```

### Comparison of estimated parameters 

```{r jSDM-boral-frogs, echo=F}
nsp <- ncol(mod_jSDM_block_Frogs$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Frogs$model_spec$presences)
nl <- mod_jSDM_block_Frogs$model_spec$n_latent
np <- nrow(mod_jSDM_block_Frogs$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,1))
plot(mod_boral_Frogs$row.coefs[[1]]$mean,summary(mod_jSDM_block_Frogs$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by boral", ylab="alphas estimated by jSDM", main="Site random effect alpha",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
points(mod_boral_Frogs$row.sigma[[1]]$mean,summary(mod_jSDM_block_Frogs$mcmc.Valpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  jSDM_betas[j,] <- summary(mod_jSDM_block_Frogs$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Frogs$lv.coefs.mean[,"beta0"],mod_boral_Frogs$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by boral", ylab="betas estimated by jSDM", main="Species fixed effect beta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  jSDM_lambdas[j,] <- summary(mod_jSDM_block_Frogs$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Frogs$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda on latent variables")
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Frogs$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Frogs$lv.mean, jSDM_lvs, xlab="latent variables estimated by boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2")
abline(a=0,b=1,col='red')

# Predictions 
plot(probit_theta_pred_Frogs, mod_jSDM_block_Frogs$probit_theta_pred, xlab="probit theta estimated by boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta")
abline(a=0,b=1,col='red')
```

## Fungi dataset : 800 sites and 11 species

### Fit with boral 
```{r boral-fungi, cache=TRUE,echo=F, include=F}

# Import center and reduce fungi dataset
Fungi <- read.csv("~/Code/data/Fungi.csv", sep=",", header=T)
Env_Fungi <- cbind(scale(Fungi[,c("diam","epi","bark")]),Fungi[,c("dc1","dc2","dc3","dc4","dc5","quality3","quality4","ground3","ground4")])
colnames(Env_Fungi) <- c("diam","epi","bark","dc1","dc2","dc3","dc4","dc5","quality3","quality4","ground3","ground4")
PA_Fungi <- Fungi[,c("antser","antsin","astfer","fompin","hetpar","junlut","phefer","phenig","phevit","poscae","triabi")]

# Fit the model 
T1 <- Sys.time()
mod_boral_Fungi <- boral(y=PA_Fungi, X=Env_Fungi, lv.control=list(num.lv=2, type="independent", distmat=NULL), family="binomial", row.eff="random", row.ids=matrix(c(1:nrow(PA_Fungi)),ncol=1), save.model=TRUE, calc.ics=F, hypparams=c(20,20,20,20), mcmc.control=list(n.burnin=30000, n.iteration=35000,n.thin=5,seed=123))
T2 <- Sys.time()
T_boral_Fungi <- difftime(T2,T1,units="mins")

# Predicted probit theta 
probit_theta_pred_Fungi=mod_boral_Fungi$row.coefs[[1]]$mean + as.matrix(Env_Fungi) %*% t(mod_boral_Fungi$X.coefs.mean)  + mod_boral_Fungi$lv.coefs.mean[,"beta0"] + mod_boral_Fungi$lv.mean%*%t(mod_boral_Fungi$lv.coefs.mean[,-1])

# Deviance
logL=0
for (i in 1:nrow(PA_Fungi)){
  for (j in 1:ncol(PA_Fungi)){
    theta <- pnorm(probit_theta_pred_Fungi[i,j])
    logL=logL + dbinom(PA_Fungi[i,j],1,theta,1)  
  }
}
Deviance_boral_Fungi <- -2*logL
```

### Fit with jSDM
```{r jSDM-fungi, cache=TRUE, echo=F, include=F}
# Fit the model
T1 <- Sys.time()
mod_jSDM_block_Fungi <- jSDM_probit_block (
  presence_site_sp=as.matrix(PA_Fungi), 
  site_suitability=~.,   
  site_data=Env_Fungi, n_latent=2,
  burnin=30000, mcmc=5000, thin=5,
  alpha_start=0, beta_start=0, lambda_start=0, W_start=0,
  V_alpha_start=1, shape=0.5, rate=0.0005,
  mu_beta=0, V_beta=1.0E6,
  mu_lambda=0, V_lambda=10,
  seed=123, verbose=1)
T2 <- Sys.time()
T_jSDM_block_Fungi <- difftime(T2,T1,units="mins")
```

### Comparison of estimated parameters 
```{r jSDM-boral-fungi, echo=F}
nsp <- ncol(mod_jSDM_block_Fungi$model_spec$presences)
nsite <- nrow(mod_jSDM_block_Fungi$model_spec$presences)
nl <- mod_jSDM_block_Fungi$model_spec$n_latent
np <- nrow(mod_jSDM_block_Fungi$model_spec$beta_start)
# Alpha et V_alpha
par(mfrow=c(1,1))
plot(mod_boral_Fungi$row.coefs[[1]]$mean,summary(mod_jSDM_block_Fungi$mcmc.alpha)[[1]][,"Mean"], xlab="alphas estimated by boral", ylab="alphas estimated by jSDM", main="Site random effect alpha",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
points(mod_boral_Fungi$row.sigma[[1]]$mean,summary(mod_jSDM_block_Fungi$mcmc.Valpha)[[1]]["Mean"],
       pch=18, col ='red',cex=2.5)
legend("bottomright", legend=c("V_alpha"), pch =18 , col ='red',pt.cex =2, cex=1.2)

# Betas
jSDM_betas <- matrix(0,nsp,np)
for (j in 1:nsp){
  jSDM_betas[j,] <- summary(mod_jSDM_block_Fungi$mcmc.sp[[paste0("sp_",j)]])[[1]][1:np,"Mean"]
}
boral_betas <- cbind(mod_boral_Fungi$lv.coefs.mean[,"beta0"],mod_boral_Fungi$X.coefs.mean)

plot(boral_betas,jSDM_betas, xlab="betas estimated by boral", ylab="betas estimated by jSDM", main="Species fixed effect beta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Lambdas
jSDM_lambdas <- matrix(0,nsp,nl)
for (j in 1:nsp){
  jSDM_lambdas[j,] <- summary(mod_jSDM_block_Fungi$mcmc.sp[[paste0("sp_",j)]])[[1]][(np+1):(np+nl),"Mean"]
}
boral_lambdas <- mod_boral_Fungi$lv.coefs.mean[,-1]

plot(boral_lambdas,jSDM_lambdas, xlab="lambdas estimated by boral", ylab="lambdas estimated by jSDM", main="Species fixed effect lambda on latent variables",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')

# Ws
jSDM_lvs <- matrix(0,nsite,nl)
for (l in 1:nl){
  jSDM_lvs[,l] <- summary(mod_jSDM_block_Fungi$mcmc.latent[[paste0("lv_",l)]])[[1]][,"Mean"]
}
plot(mod_boral_Fungi$lv.mean, jSDM_lvs, xlab="latent variables estimated by boral", ylab="latent variables estimated by jSDM", main="Latent variables W_1 and W_2",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
# Predictions 
plot(probit_theta_pred_Fungi, mod_jSDM_block_Fungi$probit_theta_pred, xlab="probit theta estimated by boral", ylab="probit theta estimated by jSDM", main="Predicted probit theta",cex.lab=1.5,cex.main=1.5,font.lab=2)
abline(a=0,b=1,col='red')
```
